Karpov, Kasprov, Anand

### 2022-04-28 inst/kka.Rmd
hyper3 likelihoods for chess with draws

@@

```{r defineH}

results <- as.list(kka)
attach(results)
H <- hyper2()

@@

```{r testall3}
samep.test(karpov_kasparov_anand,c("Karpov","Kasparov","Anand"))
```

@@

>>>>

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

@@

```{r defineH,cache=TRUE}
results <- as.list(kka)
attach(results)
H <- hyper2()

@@

```{r testall3,cache=TRUE}
samep.test(karpov_kasparov_anand,c("Karpov","Kasparov","Anand"))
```

@@

# Some ideas for `hyper3` analysis.

First we ignore the draws and create a matrix with real parts
representing white wins (home wins in the context of football) and
imaginary parts representing black (away) wins.  A pairwise likelihood function would be

$$
\left(\frac{\lambda p_1}{\lambda p_1+p_2}\right)^a
\left(\frac{        p_2}{\lambda p_1+p_2}\right)^b
\left(\frac{        p_1}{p_1+\lambda p_2}\right)^c
\left(\frac{\lambda p_1}{p_1+\lambda p_2}\right)^d
$$

for the observation of $+a\,=0\,-b$ for $p_1$ playing white and
$+c\,=0\,-d$ for $p_2$ playing white.  We then translate this into a
`hyper3` likelihood function as follows

```{r,label=kka_hyper3,cache=TRUE}
(M <- kka_array[,,1] + 1i*kka_array[,,3])
home_away3(M,lambda=1.1)
```

Above we use $\lambda=1.1$, as a specific numeric value is required by
the idiom.

```{r, showhyper3kkamaxlikeexists,cache=TRUE}
l <- function(lambda){
  H <- home_away3(M,lambda=lambda)
  loglik(maxp(H),H)
}
lam <- seq(from=1.2,to=3.5,len=10)
like <- sapply(lam,l)
```

and plot it:

```{r,label=plothyper3likecurve}
plot(lam,like-max(like),type="b",pch=16)
abline(h=c(0,-2),lty=2)
```

Above we see a credible interval for $\lambda$ being about
$(1.4,2.9)$.  Of course, this is crude and it would be betterto have
more transparent package idiom.  But I'm working on this.

We can now introduce draws but first a change of notation.  The
previous likelihood function can be represented thus:

$$
\left(\frac{\lambda p_1, p_2}{\lambda p_1+p_2}\right)^{(a,b)}
\left(\frac{p_1, \lambda p_2}{p_1+\lambda p_2}\right)^{(c,d)}
$$

Draws would change this by introducing a weighted Draw monster D$:

$$
\left(\frac{\lambda p_1, D(p_1+p_2),p_2}{\lambda p_1+D(p_1+p_2)+p_2}\right)^{(a,b,c)}
\left(\frac{p_1, D(p_1+p_2),\lambda p_2}{p_1+D(p_1+p_2)+\lambda p_2}\right)^{(d,e,f)}
$$

or possibly an individual Draw monster:

$$
\left(\frac{\lambda p_1, D_1p_1+D_2p_2,p_2}{\lambda p_1+D_1p_1+D_2p_2+p_2}\right)^{(a,b,c)}
\left(\frac{p_1, D_1p_1+D_2p_2),\lambda p_2}{p_1+D_1p_1+D_2p_2+\lambda p_2}\right)^{(d,e,f)}
$$

Note that in neither of these models do we have the probability of a
draw for $p_1$ playing white being the same as for $p_2$ playing
white.  In the first case the two probabilities are

$$
\frac{D(p_1+p_2)}{(\lambda +D)p_1 + (1+D)p_2},\qquad
\frac{D(p_1+p_2)}{(1+D)p_1 + (\lambda+D)p_2}
$$

and in the second they are

$$
\frac{D_1p_1+D_2p_2}{(\lambda +D_1)p_1 + (1+D_2)p_2},\qquad
\frac{D_1p_1+D_2p_2}{(1 +D_1)p_1 + (\lambda+D_2)p_2}
$$
